---
title: "GDSAnnotator: Whole-genome variant annotation using GDS files"
author:
- name: Xiuwen Zheng
  affiliation: Genomics Research Center, AbbVie, North Chicago, US
date: "May 2026"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{GDSAnnotator: Whole-genome variant annotation using GDS files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="##")
```

# Introduction

Variant annotation is a critical step in genomic data analysis, facilitating
the interpretation of genetic variations and their potential biological
significance. As large-scale whole-genome sequencing studies continue to
expand, efficient storage, retrieval, and manipulation of variant annotation
data have become increasingly important. The Genomic Data Structure
([GDS](https://www.bioconductor.org/packages/SeqArray)) file format offers a
scalable and storage-efficient solution for large genomic datasets, enabling
rapid access and computational efficiency.

`GDSAnnotator` is an R/Bioconductor package designed for fast and
memory-efficient annotation of variants stored in GDS files. It integrates
with Bioconductor tools and utilizes external data resources --- such as the
Ensembl Variant Effect Predictor
([VEP](https://useast.ensembl.org/info/docs/tools/vep/index.html)),
[SnpEff](https://pcingola.github.io/SnpEff/),
[gnomAD](https://gnomad.broadinstitute.org/) and
[FAVOR](https://favor.genohub.org) (Functional Annotation of Variants Online
Resource) --- to provide comprehensive functional annotations, including
predicted variant effects and functional characteristics of non-coding
variants. Compared to a plain text file, the GDS format is 30--40 times more
space-efficient when storing whole-genome annotations of FAVOR.

## Key functions

| Function            | Description                                                 |
|:--------------------|:------------------------------------------------------------|
| `seqToGDS_FAVOR`    | Convert FAVOR CSV files to GDS                              |
| `seqToGDS_gnomAD`   | Convert gnomAD VCF files to GDS                             |
| `seqToGDS_VEP`      | Convert Ensembl VEP VCF output to GDS                       |
| `seqToGDS_SnpEff`   | Convert SnpEff VCF output to GDS                            |
| `seqAnnotate`       | Annotate variants using annotations stored in GDS           |
| `seqAnnotList`      | List the annotations stored in a GDS file                   |
| `seqValueCounts`    | Calculate the counts of unique annotation values            |
| `seqUnitGroupAnnot` | Group variants based on their annotations                   |
| `seqAnnotStat`      | Calculate the summary statistics of variants & annotations  |
| `seqAnnotReport`    | Generate a summary report in HTML or Markdown format        |
| `seqAnnotGeneTable` | Per-gene annotation counts, broken down by impact           |

# Installation

`GDSAnnotator` can be installed from Bioconductor (once released):

```{r install-bioc, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("GDSAnnotator")
```

Load the package:

```{r load}
library(GDSAnnotator)
```

# Inspecting an annotation GDS file

`GDSAnnotator` is built on top of the
[SeqArray](https://www.bioconductor.org/packages/SeqArray) GDS format. Each
annotation GDS file stores per-variant annotations under
`annotation/info/<name>`. To discover which annotations are available in a
file, use `seqAnnotList()`:

```{r annot-list}
# FAVOR essential dataset, chromosome 22 (subset)
annot_gds <- system.file("extdata", "favor_chr22_sub.gds",
    package = "GDSAnnotator")

ann <- seqAnnotList(annot_gds)
ann
```

The returned `DataFrame` has four columns: the annotation `name`, the R data
`type`, the GDS storage `trait`, and a textual `description` (when
available).

# Annotating variants

The `seqAnnotate()` generic accepts variants in several forms: a character
vector of `chr-pos-ref-alt` strings, a `data.frame`/`DataFrame` with
chromosome, position, reference and alternative allele columns, a `GRanges` or
`GRangesList`, or an open `SeqVarGDSClass` object.

## From a character vector of variant IDs

```{r annot-char}
## Variant IDs are split on `-`, `_` or `:` (the default `split` pattern),
## so all of these formats are accepted:
##   "22-10510007-T-C"  "22:10510007:T:C"  "22_10510007_T_C"
snps <- c(
    "22-10510007-T-C",
    "22-10510038-T-C",
    "22-10510282-G-C",
    "22-10510303-G-C"   # a fake SNP, returns NA
)

# Use all annotations available in the file
varnm <- ann$name
seqAnnotate(snps, annot_gds, varnm)
```

A subset of annotations can be selected by passing only the names of
interest in `varnm`:

```{r annot-subset}
seqAnnotate(snps, annot_gds,
    varnm = c("apc_conservation", "cadd_phred", "linsight"))
```

## From a `data.frame`

```{r annot-df}
df <- data.frame(
    chr = c("22", "22", "22"),
    pos = c(10510007L, 10510038L, 10510282L),
    ref = c("T", "T", "G"),
    alt = c("C", "C", "C")
)
df

seqAnnotate(df, annot_gds, varnm = c("apc_conservation", "cadd_phred"))
```

## From a `GRanges` object

`GRanges` queries return all variants that fall within the specified ranges:

```{r annot-granges}
gr <- GenomicRanges::GRanges(
    seqnames = c("chr22", "chr22"),
    ranges = IRanges::IRanges(
        start = c(10510100, 10510225),
        end   = c(10510125, 10510250)))

seqAnnotate(gr, annot_gds, varnm = c("apc_conservation", "cadd_phred"))
```

## Writing annotations into a SeqArray GDS file

When the object being annotated is a `SeqVarGDSClass` genotype GDS file,
annotations can be written directly into the same file (under
`annotation/info/`) by setting `add_to_gds = TRUE`, or into a new GDS file by
passing a file name. This avoids materialising a large annotation
`DataFrame` in R memory for whole-genome studies.

```{r annot-gds, eval = FALSE}
library(SeqArray)

# An existing genotype GDS file
geno_gds <- "my_wgs_geno.gds"

seqAnnotate(seqOpen(geno_gds), annot_gds, varnm,
    add_to_gds = "my_wgs_geno_annotated.gds")
```

# Converting annotation sources to GDS

`GDSAnnotator` provides dedicated importers for common annotation sources.
All of them output a SeqArray-compatible GDS file, which can subsequently be
queried with `seqAnnotate()`.

## FAVOR

Convert FAVOR CSV files (one chromosome per file) to GDS:

```{r conv-favor}
csv_fn <- system.file("extdata", "favor_chr22_sub.csv.gz",
    package = "GDSAnnotator")

out_fn <- tempfile(fileext = ".gds")
seqToGDS_FAVOR(csv_fn, out_fn, verbose = FALSE)

unlink(out_fn, force = TRUE)
```

When FAVOR is distributed as a tar archive of per-chromosome CSVs,
`seqToGDS_FAVOR_tar()` can be used to convert directly from the tar file.

## gnomAD

```{r conv-gnomad}
vcf_fn <- system.file("extdata", "gnomad.genomes.v4.chr22_sub.vcf.gz",
    package = "GDSAnnotator")

out_fn <- tempfile(fileext = ".gds")
seqToGDS_gnomAD(vcf_fn, out_fn, verbose = FALSE)

unlink(out_fn, force = TRUE)
```

## Ensembl VEP

The VEP importer parses the compound `CSQ` INFO field emitted by Ensembl VEP
into individual annotation fields in the GDS file.

```{r conv-vep}
vcf_fn <- system.file("extdata", "example_wgs_sites_chr22_vep.vcf.gz",
    package = "GDSAnnotator")

out_fn <- tempfile(fileext = ".gds")
seqToGDS_VEP(vcf_fn, out_fn, verbose = FALSE)

unlink(out_fn, force = TRUE)
```

## SnpEff

The SnpEff importer parses the `ANN`, `LOF` and `NMD` INFO fields by
default:

```{r conv-snpeff}
vcf_fn <- system.file("extdata", "example_wgs_sites_chr22_snpeff.vcf.gz",
    package = "GDSAnnotator")

out_fn <- tempfile(fileext = ".gds")
seqToGDS_SnpEff(vcf_fn, out_fn, verbose = FALSE)

unlink(out_fn, force = TRUE)
```

The resulting GDS file can be inspected with `seqAnnotList()` and queried with
`seqAnnotate()` just like the FAVOR example above.

The two files converted above are also shipped with the package, as
`example_wgs_sites_chr22_vep.gds` and `example_wgs_sites_chr22_snpeff.gds`,
and are used in the examples that follow.

# Summarising annotations

## Counting unique values

`seqValueCounts()` is a convenience wrapper that tabulates unique values of
one or more annotation fields without loading the entire annotation into R:

```{r value-counts}
gds_fn <- system.file("extdata", "gnomad.genomes.v4.chr22_sub.gds",
    package = "GDSAnnotator")

f <- SeqArray::seqOpen(gds_fn)

seqValueCounts(f, "variant_type")
seqValueCounts(f, c("variant_type", "allele_type"))

SeqArray::seqClose(f)
```

Set `per_variant = TRUE` to collapse multi-valued annotations (e.g. multiple
VEP consequences per variant) to a single count per variant, and
`parallel = TRUE` to enable multi-core processing for large files.

## Grouping variants by annotation

`seqUnitGroupAnnot()` groups variants by the values of one or more
annotation fields, optionally applying filters via `cond`. This is useful
for constructing variant-set units for downstream aggregation tests (e.g.
rare-variant association analysis):

```{r unit-group}
gds_fn <- system.file("extdata", "example_wgs_sites_chr22_snpeff.gds",
    package = "GDSAnnotator")

# Group the variants by gene, keeping only the HIGH-impact annotations
u <- seqUnitGroupAnnot(gds_fn,
    varnm = c("ANN.list/Gene_Name", "ANN.list/Annotation_Impact"),
    by    = "ANN.list/Gene_Name",
    cond  = list(NULL, "HIGH"),
    verbose = FALSE)

length(u$index)   # the number of variant-set units
head(u$desp)
```

## Summary report

Ensembl VEP and SnpEff each write an HTML summary next to their VCF output,
but that report is lost once the annotations have been converted to GDS ---
and it is not available at all for a GDS file assembled from several sources.
`seqAnnotStat()` and `seqAnnotReport()` regenerate an equivalent report
directly from the GDS file.

`seqAnnotStat()` collects every statistic in a **single pass** over the file,
so the cost does not grow with the number of statistics reported:

```{r annot-stat}
gds_fn <- system.file("extdata", "example_wgs_sites_chr22_snpeff.gds",
    package = "GDSAnnotator")

st <- seqAnnotStat(gds_fn, verbose = FALSE)
st
```

Each statistic in `st$counts` is a named vector of counts. Consequences are
counted twice: once for every annotation record (one record per transcript,
`cons_all`), and once per variant keeping only the most severe consequence
(`cons_severe`), following the Sequence Ontology ranking used by VEP:

```{r annot-stat-counts}
st$counts$cons_severe

st$counts$impact
```

`seqAnnotGeneTable()` returns the per-gene counts, the equivalent of the
`genes.txt` file written by SnpEff:

```{r annot-gene-table}
head(seqAnnotGeneTable(st))
```

`seqAnnotReport()` renders the statistics as a report. The output format is
taken from the file extension by default: `.md` gives GitHub-flavored
Markdown, `.Rmd` gives R Markdown source, and anything else gives a
self-contained HTML file that needs no JavaScript and no additional package,
so it can be viewed offline. None of the three writes a companion image
directory --- the report is always a single file.

```{r annot-report-md}
md_fn <- tempfile(fileext = ".md")
seqAnnotReport(st, md_fn, title = "SnpEff annotation of chromosome 22",
    verbose = FALSE)

cat(head(readLines(md_fn), 22), sep = "\n")
```

```{r annot-report-html, eval = FALSE}
# the same content as a self-contained HTML file
seqAnnotReport(st, "annot_summary.html",
    title = "SnpEff annotation of chromosome 22")
```

The HTML charts are inline SVG and the Markdown charts are drawn with block
characters. If the charts should be real plots instead --- to restyle them,
or to fold the report into a larger document --- the `rmd` format writes an R
Markdown source in which the counts are embedded as R code and the charts are
produced by `barplot()` at knit time:

```{r annot-report-rmd, eval = FALSE}
seqAnnotReport(st, "annot_summary.Rmd",
    title = "SnpEff annotation of chromosome 22")

rmarkdown::render("annot_summary.Rmd")
```

```{r annot-report-clean, include = FALSE}
unlink(md_fn, force = TRUE)
```

The counters are additive, and that has two consequences. First, the variants
can be split between processes and the per-process counters merged
afterwards, so `parallel` gives a summary identical to the serial one:

```{r annot-stat-parallel, eval = FALSE}
st <- seqAnnotStat(gds_fn, parallel = 4L)
```

Second, a whole-genome summary can be built from per-chromosome GDS files
without ever holding more than one of them open. The memory usage depends on
the number of distinct categories (e.g. the number of genes), not on the
number of variants:

```{r annot-stat-merge, eval = FALSE}
gds_fn <- sprintf("annot_chr%d.gds", 1:22)

st <- seqAnnotStat(gds_fn)
seqAnnotReport(st, "wgs_summary.html", title = "Whole-genome annotation")
```

# Performance: compact storage and random access

Two properties make the GDS representation attractive for whole-genome
annotation. First, the columnar, typed and block-compressed layout is far more
compact than a flat text table. Second --- and unlike a gzip-compressed text
file, which must be decompressed and scanned end-to-end for every lookup --- a
GDS file supports *random access*: `seqAnnotate()` reads only the variants and
the annotation columns it actually needs, so memory use stays flat regardless
of how large the underlying database is.

```{r perf-size}
annot_gds <- system.file("extdata", "favor_chr22_sub.gds",
    package = "GDSAnnotator")
ann <- seqAnnotList(annot_gds)

# materialise the same content as a flat text table for comparison
f <- SeqArray::seqOpen(annot_gds)
tab <- data.frame(
    chr = seqGetData(f, "$chromosome"), pos = seqGetData(f, "position"),
    ref = seqGetData(f, "$ref"),        alt = seqGetData(f, "$alt"))
for (nm in ann$name)
    tab[[nm]] <- seqGetData(f, paste0("annotation/info/", nm), .tolist = NA)
SeqArray::seqClose(f)

tsv <- tempfile(fileext = ".tsv")
write.table(tab, tsv, sep = "\t", row.names = FALSE, quote = FALSE)
raw <- readBin(tsv, "raw", file.size(tsv))

sizes <- c(
    `GDS (LZMA)`     = file.size(annot_gds),
    `flat text`      = file.size(tsv),
    `flat text+gzip` = length(memCompress(raw, "gzip")))
data.frame(KB = round(sizes / 1024, 1),
    `x vs GDS` = round(sizes / sizes[["GDS (LZMA)"]], 1),
    check.names = FALSE)
unlink(tsv, force = TRUE)
```

```{r perf-query}
# random-access query latency (only the requested variants are decoded)
set.seed(1)
k   <- sample(nrow(tab), 100)
qry <- paste(tab$chr[k], tab$pos[k], tab$ref[k], tab$alt[k], sep = "-")

system.time(for (i in 1:5)
    seqAnnotate(qry, annot_gds, varnm = ann$name, verbose = FALSE))
```

On this 1,000-variant subset the GDS file is already an order of magnitude
smaller than the raw text table, and about the same size as gzip-compressed
text --- but with the random access that gzip cannot offer. On the complete
whole-genome FAVOR database the ratio against raw text reaches 30--40x, while
per-query latency and memory stay essentially constant because only the
requested variants are read. A script for benchmarking at that scale ships
with the package:

```{r perf-script, eval = FALSE}
system.file("benchmark", "benchmark.R", package = "GDSAnnotator")
```

# Session information

```{r session-info}
sessionInfo()
```
