---
title: "GDSAnnotator: Whole-genome variant annotation using GDS files"
author:
- name: Xiuwen Zheng
  affiliation: Genomics Research Center, AbbVie, North Chicago, US
date: "May 2026"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{GDSAnnotator: Whole-genome variant annotation using GDS files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="##")
```

# Introduction

Variant annotation is a critical step in genomic data analysis, facilitating
the interpretation of genetic variations and their potential biological
significance. As large-scale whole-genome sequencing studies continue to
expand, efficient storage, retrieval, and manipulation of variant annotation
data have become increasingly important. The Genomic Data Structure
([GDS](https://www.bioconductor.org/packages/SeqArray)) file format offers a
scalable and storage-efficient solution for large genomic datasets, enabling
rapid access and computational efficiency.

`GDSAnnotator` is an R/Bioconductor package designed for fast and
memory-efficient annotation of variants stored in GDS files. It integrates
with Bioconductor tools and utilizes external data resources --- such as the
Ensembl Variant Effect Predictor
([VEP](https://useast.ensembl.org/info/docs/tools/vep/index.html)),
[SnpEff](https://pcingola.github.io/SnpEff/),
[gnomAD](https://gnomad.broadinstitute.org/) and
[FAVOR](https://favor.genohub.org) (Functional Annotation of Variants Online
Resource) --- to provide comprehensive functional annotations, including
predicted variant effects and functional characteristics of non-coding
variants. Compared to a plain text file, the GDS format is 30--40 times more
space-efficient when storing whole-genome annotations of FAVOR.

## Key functions

| Function            | Description                                                 |
|:--------------------|:------------------------------------------------------------|
| `seqToGDS_FAVOR`    | Convert FAVOR CSV files to GDS                              |
| `seqToGDS_gnomAD`   | Convert gnomAD VCF files to GDS                             |
| `seqToGDS_VEP`      | Convert Ensembl VEP VCF output to GDS                       |
| `seqToGDS_SnpEff`   | Convert SnpEff VCF output to GDS                            |
| `seqAnnotate`       | Annotate variants using annotations stored in GDS           |
| `seqAnnotList`      | List the annotations stored in a GDS file                   |
| `seqValueCounts`    | Calculate the counts of unique annotation values            |
| `seqUnitGroupAnnot` | Group variants based on their annotations                   |

# Installation

`GDSAnnotator` can be installed from Bioconductor (once released):

```{r install-bioc, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("GDSAnnotator")
```

Load the package:

```{r load}
library(GDSAnnotator)
```

# Inspecting an annotation GDS file

`GDSAnnotator` is built on top of the
[SeqArray](https://www.bioconductor.org/packages/SeqArray) GDS format. Each
annotation GDS file stores per-variant annotations under
`annotation/info/<name>`. To discover which annotations are available in a
file, use `seqAnnotList()`:

```{r annot-list}
# FAVOR essential dataset, chromosome 22 (subset)
annot_gds <- system.file("extdata", "favor_chr22_sub.gds",
    package = "GDSAnnotator")

ann <- seqAnnotList(annot_gds)
ann
```

The returned `DataFrame` has four columns: the annotation `name`, the R data
`type`, the GDS storage `trait`, and a textual `description` (when
available).

# Annotating variants

The `seqAnnotate()` generic accepts variants in several forms: a character
vector of `chr-pos-ref-alt` strings, a `data.frame`/`DataFrame` with
chromosome, position, reference and alternative allele columns, a `GRanges` or
`GRangesList`, or an open `SeqVarGDSClass` object.

## From a character vector of variant IDs

```{r annot-char}
## Variant IDs are split on `-`, `_` or `:` (the default `split` pattern),
## so all of these formats are accepted:
##   "22-10510007-T-C"  "22:10510007:T:C"  "22_10510007_T_C"
snps <- c(
    "22-10510007-T-C",
    "22-10510038-T-C",
    "22-10510282-G-C",
    "22-10510303-G-C"   # a fake SNP, returns NA
)

# Use all annotations available in the file
varnm <- ann$name
seqAnnotate(snps, annot_gds, varnm)
```

A subset of annotations can be selected by passing only the names of
interest in `varnm`:

```{r annot-subset}
seqAnnotate(snps, annot_gds,
    varnm = c("apc_conservation", "cadd_phred", "linsight"))
```

## From a `data.frame`

```{r annot-df}
df <- data.frame(
    chr = c("22", "22", "22"),
    pos = c(10510007L, 10510038L, 10510282L),
    ref = c("T", "T", "G"),
    alt = c("C", "C", "C")
)
df

seqAnnotate(df, annot_gds, varnm = c("apc_conservation", "cadd_phred"))
```

## From a `GRanges` object

`GRanges` queries return all variants that fall within the specified ranges:

```{r annot-granges}
gr <- GenomicRanges::GRanges(
    seqnames = c("chr22", "chr22"),
    ranges = IRanges::IRanges(
        start = c(10510100, 10510225),
        end   = c(10510125, 10510250)))

seqAnnotate(gr, annot_gds, varnm = c("apc_conservation", "cadd_phred"))
```

## Writing annotations into a SeqArray GDS file

When the object being annotated is a `SeqVarGDSClass` genotype GDS file,
annotations can be written directly into the same file (under
`annotation/info/`) by setting `add_to_gds = TRUE`, or into a new GDS file by
passing a file name. This avoids materialising a large annotation
`DataFrame` in R memory for whole-genome studies.

```{r annot-gds, eval = FALSE}
library(SeqArray)

# An existing genotype GDS file
geno_gds <- "my_wgs_geno.gds"

seqAnnotate(seqOpen(geno_gds), annot_gds, varnm,
    add_to_gds = "my_wgs_geno_annotated.gds")
```

# Converting annotation sources to GDS

`GDSAnnotator` provides dedicated importers for common annotation sources.
All of them output a SeqArray-compatible GDS file, which can subsequently be
queried with `seqAnnotate()`.

## FAVOR

Convert FAVOR CSV files (one chromosome per file) to GDS:

```{r conv-favor}
csv_fn <- system.file("extdata", "favor_chr22_sub.csv.gz",
    package = "GDSAnnotator")

out_fn <- tempfile(fileext = ".gds")
seqToGDS_FAVOR(csv_fn, out_fn, verbose = FALSE)

unlink(out_fn, force = TRUE)
```

When FAVOR is distributed as a tar archive of per-chromosome CSVs,
`seqToGDS_FAVOR_tar()` can be used to convert directly from the tar file.

## gnomAD

```{r conv-gnomad}
vcf_fn <- system.file("extdata", "gnomad.genomes.v4.chr22_sub.vcf.gz",
    package = "GDSAnnotator")

out_fn <- tempfile(fileext = ".gds")
seqToGDS_gnomAD(vcf_fn, out_fn, verbose = FALSE)

unlink(out_fn, force = TRUE)
```

## Ensembl VEP

The VEP importer parses the compound `CSQ` INFO field emitted by Ensembl VEP
into individual annotation fields in the GDS file.

```{r conv-vep}
vcf_fn <- system.file("extdata", "example_wgs_sites_chr22_vep.vcf.gz",
    package = "GDSAnnotator")

out_fn <- tempfile(fileext = ".gds")
seqToGDS_VEP(vcf_fn, out_fn, verbose = FALSE)

unlink(out_fn, force = TRUE)
```

## SnpEff

The SnpEff importer parses the `ANN`, `LOF` and `NMD` INFO fields by
default:

```{r conv-snpeff}
vcf_fn <- system.file("extdata", "example_wgs_sites_chr22_snpeff.vcf.gz",
    package = "GDSAnnotator")

out_fn <- tempfile(fileext = ".gds")
seqToGDS_SnpEff(vcf_fn, out_fn, verbose = FALSE)

unlink(out_fn, force = TRUE)
```

The resulting GDS file can be inspected with `seqAnnotList()` and queried with
`seqAnnotate()` just like the FAVOR example above.

# Summarising annotations

## Counting unique values

`seqValueCounts()` is a convenience wrapper that tabulates unique values of
one or more annotation fields without loading the entire annotation into R:

```{r value-counts}
gds_fn <- system.file("extdata", "gnomad.genomes.v4.chr22_sub.gds",
    package = "GDSAnnotator")

f <- SeqArray::seqOpen(gds_fn)

seqValueCounts(f, "variant_type")
seqValueCounts(f, c("variant_type", "allele_type"))

SeqArray::seqClose(f)
```

Set `per_variant = TRUE` to collapse multi-valued annotations (e.g. multiple
VEP consequences per variant) to a single count per variant, and
`parallel = TRUE` to enable multi-core processing for large files.

## Grouping variants by annotation

`seqUnitGroupAnnot()` groups variants by the values of one or more
annotation fields, optionally applying filters via `cond`. This is useful
for constructing variant-set units for downstream aggregation tests (e.g.
rare-variant association analysis):

```{r unit-group, eval = FALSE}
gds_fn <- system.file("extdata", "example_wgs_sites_chr22_vep.gds",
    package = "GDSAnnotator")

# Group variants by VEP 'Consequence', keeping only coding consequences
seqUnitGroupAnnot(gds_fn,
    varnm = c("CSQ_Consequence", "CSQ_IMPACT"),
    by    = "CSQ_IMPACT",
    cond  = list(NULL, c("HIGH", "MODERATE")))
```

# Session information

```{r session-info}
sessionInfo()
```
